---
title: "Reproducible Research: Peer Assessment 1"
author: Angela Liu
date: Sept. 12, 2018 
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r read_in_data}

data <- read.csv(file="activity.csv", header=TRUE, sep=",")

dim(data)
head(data)
str(data)

```

## What is the mean total number of steps taken per day?

### First, compute each day's total value of steps. 
```{r per-day-sum}

library('plyr') # load plyr package

per_day_total <- ddply(data, .(date), summarize, sum=sum(steps)) 

head(per_day_total)

```

### Then, plot a histogram of the total steps per day. 
```{r hist-per-day, fig.height=5}

hist(per_day_total$sum, nclass=30, main="Histogram of total number of steps per day", xlab="Total number of steps per day", ylab="Count")


```

### Finally, compute the mean and median of the above total steps. 
```{r per-day-stat}

summary(per_day_total$sum) 

```


## What is the average daily activity pattern?

```{r per-interval-mean, fig.height=5}

na_omit_data = na.omit(data)

per_interval_stat <- ddply(na_omit_data, .(interval), summarize, sum=sum(steps), means=mean(steps)) 

plot(per_interval_stat$interval, per_interval_stat$means, type="l", main="Average number of steps per interval over time", xlab="Time interval in 5-minute", ylab="Average number of steps per interval")

```


### Find the time interval that has the maximum average number of steps across days. 


```{r per_interval_max}

index = which.max(per_interval_stat$means)

per_interval_stat$interval[index]


```


## Imputing missing values

### First find the number of rows with NA or missing values.  
```{r imputing_NA_rows}
num_rows_NA <- sum(is.na(data$steps))

```

There are `r num_rows_NA` rows with missing values or "NA" in the original data. 


### We will use the average number of steps across all of the rest of days with real data per every 5-minute interval to replace all of the "NA" for the corresponding interval. 

```{r imputing_NA_mean}
imputed_data <- data

index_NA_rows <- which(is.na(data$steps), arr.ind=TRUE)

# try a for loop
for (index in index_NA_rows) {
    interval = imputed_data$interval[index]
    imputed_data$steps[index] = per_interval_stat$means[ per_interval_stat$interval == interval ]
}

# the below method only worked for the first 288 rows of data
#imputed_data$steps[index_NA_rows] = per_interval_stat$means[(imputed_data$interval[index_NA_rows] == per_interval_stat$interval)]

head(imputed_data)
tail(imputed_data)

```

### Plot a new histogram using the imputed data. 

```{r impute_hist}
imputed_per_day_total <- ddply(imputed_data, .(date), summarize, sum=sum(steps)) 

hist(imputed_per_day_total$sum, nclass=30, main="Histogram of total number of steps per day (Imputed)", xlab="Total number of steps per day", ylab="Count")
```


### Compute the mean and median of daily number of steps with imputed data.  

```{r impute_stat}
summary(imputed_per_day_total$sum) 
```

As shown by the new histogram plot and new summary statistics, by imputing the missing values with average number of steps per 5-minute interval, the new histogram has a significantly higher peak around 10766, which is the mean value of the daily number of steps.  


## Are there differences in activity patterns between weekdays and weekends?

### First, we will create a new vector, which is a factor with two levels, "weekday" and "weekend". 

```{r day_of_week}

day_status = weekdays(as.Date(data$date))
day_status[day_status != "Saturday" & day_status != "Sunday"] = "weekday"
day_status[day_status == "Saturday" | day_status == "Sunday"] = "weekend"
sum(day_status == "weekend")
sum(day_status == "weekday")

```


### Then, we will add this vector as a new column to the imputed data.

```{r new_data}
new_data = cbind(imputed_data, day_status)
head(new_data)

```

### Finally, we will create a multi-panel plot showing the per-5-minute interval average number of steps across all days, split into "weekend" view and "weekday" view. 

```{r weekday_view, fig.height=8}

weekday_data = new_data[new_data$day_status == "weekday", ]
weekend_data = new_data[new_data$day_status == "weekend", ]

weekday_interval_stat <- ddply(weekday_data, .(interval), summarize, sum=sum(steps), means=mean(steps)) 
weekend_interval_stat <- ddply(weekend_data, .(interval), summarize, sum=sum(steps), means=mean(steps)) 

par(mfrow=c(2,1), oma=c(1, 0, 0, 0))
plot(weekday_interval_stat$interval, weekday_interval_stat$means, type='l', main="weekday", xlab="", ylab="")
plot(weekend_interval_stat$interval, weekend_interval_stat$means, type='l', main="weekend", xlab="", ylab="")

mtext("Time Interval", side=1, outer=TRUE, cex=1.5)
mtext("Average number of steps over multiple days", side=2, outer=TRUE, line=-1, cex=1.5)
  

```


## Render Rmd file to HTML

```{r render}

# There appears to be a bug with knitr and the first label in an R chunk will get a
#     duplicate label error. E.g., see: 
#     https://stackoverflow.com/questions/36868287/purl-within-knit-duplicate-label-error/49286316
# I have chosen to ignore this bug. 

library(knitr)
knit("PA1_template.Rmd", "PA1_template.md")
rmarkdown::render("PA1_template.md", output_file="PA1_template.html")
browseURL("PA1_template.html")

```

